---
title: "Analysis Res0sem Microbiome Data"
author: Natacha Bodenhausen
date: '`r Sys.Date()`'
output:
  pdf_document:
        toc: yes
urlcolor: blue
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)

suppressPackageStartupMessages(library(tidyverse))

library("phyloseq")
library("vegan") # it's included in phyloseq

library(cowplot) # plot_grid
library(lmerTest)  # automatically loads lme4 too
library(emmeans)  # for posthoc test
library("RColorBrewer")



set.seed(8037)



```


```{r}
my_theme <- theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

  
theme_main <-theme(text = element_text(size = 10))

custom_colors <- c("alternative" = "#E76BF3", "ThermoSeed"= "#F8766D", "synthetic" = "#00BF7D", "Tillecur" = "#A3A500", "Evonta" = "#00B0F6", "soil" = "brown", "control" = "grey")


makeFig <- TRUE
```

\newpage



```{r, eval = TRUE}
sample_table_ITS <- read_csv("../final_data/sample_ITS.csv")


sample_table_ITS <- sample_table_ITS %>% column_to_rownames("SampleID")


sample_table_16S <- read_csv("../final_data/sample_16S.csv")


sample_table_16S <- sample_table_16S %>% column_to_rownames("SampleID")



```



```{r, eval = TRUE}
tax_ITS<- read.csv("../final_data/tax_ITS.csv")

tax_ITS <-as.matrix(tax_ITS)



otu_ITS <- read.csv("../final_data/otu_ITS.csv")

otu_ITS <-as.matrix(otu_ITS)


#otu_ITS <- as.matrix(otu_ITS)


physeq_ITS <- phyloseq(otu_table(otu_ITS, taxa_are_rows = TRUE), 
                       tax_table(tax_ITS),
                       sample_data(sample_table_ITS))


```




```{r, eval = TRUE}
tax_16S<- read.csv("../final_data/tax_16S.csv")

tax_16S <-as.matrix(tax_16S)



otu_16S <- read.csv("../final_data/otu_16S.csv")

otu_16S <-as.matrix(otu_16S)


#otu_ITS <- as.matrix(otu_ITS)


physeq_16S <- phyloseq(otu_table(otu_16S, taxa_are_rows = TRUE), 
                       tax_table(tax_16S),
                       sample_data(sample_table_16S))


```


# Quality of sequencing

## remove non Fungi (wrong assignement)


```{r}
# in the new dataset, 
# there are no Annelida 

# physeq_ITS <- subset_taxa(physeq_ITS, !Phylum=="Annelida"  | is.na(Phylum))

physeq_ITS <- subset_taxa(physeq_ITS, !Phylum=="Cercozoa"| is.na(Phylum))

```

## remove  mitocondria  and chloroplast


```{r}

# Mitochondria are at the phylum level
physeq_16S <-  subset_taxa( physeq_16S,
                        !Family== "Mitochondria"| is.na(Family))

# Chloroplast are at the order level
physeq_16S <-  subset_taxa( physeq_16S,
                        !Order== "Chloroplast"| is.na(Order))

```


## ITS rarefaction plot

```{r prepare_rarefaction_plot, include=FALSE}

for_vegan_ITS <-t(as.data.frame(otu_table(physeq_ITS, 
                                                 taxa_are_rows =TRUE)))

# default plot is all in black, with black lines for the number of samples
rarefaction_ITS <- rarecurve(for_vegan_ITS, 
                                step = 1000, 
                                sample = 10000, 
                                xlab="number of sequences",
                                ylab="number of OTUs",label=FALSE)

```


```{r}
plot_data_ITS <- rarefaction_ITS %>% 
  purrr::map(~ tibble(
    y = as.vector(.x),
    x = unname(attr(.x, "Subsample")),
    steps = names(.x))) %>%
  tibble(data = ., sample = seq_along(.)) %>%
  unnest(data)

stopifnot(identical(sort(unique(plot_data_ITS$sample)), 1:NROW(sample_data(physeq_ITS))))
```


```{r, eval = TRUE}
number_sequences_perSample_ITS <-sample_sums(physeq_ITS)

#min_number <-min(number_sequences_perSample_ITS)
raremax_ITS <- 5000

## color by habitat
plot_data_ITS$color <- sample_data(physeq_ITS)$habitat [plot_data_ITS$sample]


rare_ITS <- ggplot(plot_data_ITS) +
  geom_line(aes(x , y, color = color, 
                group = sample)) +
  geom_vline(xintercept = raremax_ITS, 
             color = "darkgrey") +
  #scale_color_manual(values = six_colors) +
  labs(title = "Fungi",
       x = "number of sequences", 
       y = "number of OTUs", 
       color = NULL) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))+
  theme(legend.position = "bottom")

print(rare_ITS)






```

\newpage

## ITS transform data

```{r, echo=TRUE}

physeq_ITS_norm<-  transform_sample_counts(physeq_ITS, function(x) x / sum(x) )


physeq_ITS_hell<-  transform_sample_counts(physeq_ITS, function(x) sqrt(x / sum(x) ))


```

## ITS Stacked barplot at the phylum level

Focus on the control samples samples 

```{r}
physeq_ITS_control <- subset_samples(physeq_ITS_norm, 
                                  traitement =="control")

physeq_ITS_control_norm<-  transform_sample_counts(physeq_ITS_control, function(x) x / sum(x)*100 )

##colSums(otu_table(physeq_ITS_control_norm))

## remove OTU which are not in the control

physeq_ITS_control_norm <- prune_taxa(
                                   rowSums(otu_table(physeq_ITS_control_norm))>0, physeq_ITS_control_norm)


```

```{r}


# aggregate the OTUs with same Phylum names
dat_merge_ITS <- tax_glom(physeq = physeq_ITS_control_norm, 
                      "Phylum", NArm = FALSE) 
#colSums(otu_table(dat_merge))

molten_dat_phylum_merge_ITS <- psmelt(dat_merge_ITS)

```

```{r}

molten_dat_phylum_merge_ITS<- molten_dat_phylum_merge_ITS %>%
  mutate("Taxonomy" = case_when(Abundance >= 2 ~ Phylum,
                             Abundance < 2 ~ "less than 2%"))%>% 
  mutate(Taxonomy = replace_na(Taxonomy,"not assigned"))

```


```{r}
### Defining  colors by phylum

# display.brewer.pal(9, name = "Set1")
#  brewer.pal(9, name = "Set1")
# 
# 


#levels(as.factor(molten_dat_phylum_merge_ITS$Taxonomy))

plotColorBarplot_ITS=c(
    "Ascomycota"="#E41A1C",
    "Basidiomycota"="#377EB8",
    "Chytridiomycota"="#FF7F00",
    #"Glomeromycota"="#999999",
    #"Monoblepharomycota"= "#FF7F00",
    "Mortierellomycota"= "#F781BF",
    #"Olpidiomycota" = "#FFFF33",
    "Rozellomycota" = "#984EA3",
    "Zoopagomycota" = "#A65628",
    "less than 2%"="grey50",
    "not assigned" = "grey90")

## Same colors as Bodenhausen et al 2023
```

```{r}
molten_dat_phylum_merge_ITS %>% 
  group_by(Phylum)%>% 
  summarise(mean_value = mean(Abundance)) %>%
  arrange(desc(mean_value))
```


```{r}


p_phyla_ITS <- ggplot(molten_dat_phylum_merge_ITS, 
                  aes(x = code, 
                             y = Abundance,
                             fill = Taxonomy)) + 
  geom_bar(stat = "identity", position = "stack")+
  facet_grid(habitat  ~ management, scales = "free_x", space = "free_x")+
   scale_fill_manual(values = plotColorBarplot_ITS)+
  labs( title = "Fungi",
        y = "Relative abundance (%)", 
        x = NULL)+
  theme(legend.position="bottom")+
  theme(axis.text.x = element_text(size = 6))
  
  

p_phyla_ITS

```




## 16S rarefaction plot

```{r prepare_rarefaction_16S, include=FALSE}

for_vegan_16S <-t(as.data.frame(otu_table(physeq_16S, 
                                                 taxa_are_rows =TRUE)))

# default plot is all in black, with black lines for the number of samples
rarefaction_16S <- rarecurve(for_vegan_16S, 
                                step = 1000, 
                                sample = 10000, 
                                xlab="number of sequences",
                                ylab="number of OTUs",label=FALSE)

```


```{r}
plot_data_16S <- rarefaction_16S %>% 
  purrr::map(~ tibble(
    y = as.vector(.x),
    x = unname(attr(.x, "Subsample")),
    steps = names(.x))) %>%
  tibble(data = ., sample = seq_along(.)) %>%
  unnest(data)

stopifnot(identical(sort(unique(plot_data_16S$sample)), 1:NROW(sample_data(physeq_16S))))
```


```{r, eval = TRUE}
number_sequences_perSample_16S <-sample_sums(physeq_16S)

min_number <-min(number_sequences_perSample_16S)

## color by habitat
plot_data_16S$color <- sample_data(physeq_16S)$habitat [plot_data_16S$sample]


rare_16S <- ggplot(plot_data_16S) +
  geom_line(aes(x , y, color = color, 
                group = sample)) +
  geom_vline(xintercept = min_number, 
             color = "darkgrey") +
  #scale_color_manual(values = six_colors) +
  labs(title = "Bacteria",
       x = "number of sequences", 
       y = "number of OTUs", 
       color = NULL) +
  theme_bw() +
   theme(plot.title = element_text(hjust = 0.5))+
  theme(legend.position = "none")


print(rare_16S)






```

\newpage

## 16S transform data

```{r, echo=TRUE}

physeq_16S_norm<-  transform_sample_counts(physeq_16S, function(x) x / sum(x) )


physeq_16S_hell<-  transform_sample_counts(physeq_16S, function(x) sqrt(x / sum(x) ))


```

## 16S Stacked barplot at the phylum level

Focus on the control samples samples  (soil and rhizosphere)

```{r}
physeq_16S_control <- subset_samples(physeq_16S_norm, 
                                  traitement =="control")

physeq_16S_control_norm<-  transform_sample_counts(physeq_16S_control, function(x) x / sum(x)*100 )

##colSums(otu_table(physeq_16S_control_norm))

## remove OTU which are not in the control

physeq_16S_control_norm <- prune_taxa(
                                   rowSums(otu_table(physeq_16S_control_norm))>0, physeq_16S_control_norm)


```

```{r}


# aggregate the OTUs with same Phylum names
dat_merge_16S <- tax_glom(physeq = physeq_16S_control_norm, 
                      "Phylum", NArm = FALSE) 
#colSums(otu_table(dat_merge))

molten_dat_phylum_merge_16S <- psmelt(dat_merge_16S)

```

```{r}

# the NA not assigned are less than 3%
# molten_dat_phylum_merge_16S %>% 
#   filter(is.na(Phylum)) %>% 
#   select(Abundance)


molten_dat_phylum_merge_16S<- molten_dat_phylum_merge_16S %>%
  mutate("Taxonomy" = case_when(Abundance >= 2.5 ~ Phylum,
                             Abundance < 2.5 ~ "less than 2.5%"))%>% 
  mutate(Taxonomy = replace_na(Taxonomy,"not assigned"))

```


```{r}
### Defining  colors by phylum

# display.brewer.pal(n=12,
#                      name = "Set3")
# 
 brewer.pal(12, name = "Set3")

# levels(as.factor(molten_dat_phylum_merge_16S$Taxonomy))

plotColorBarplot_16S=c(
    "Acidobacteriota"="#8DD3C7",
    "Actinobacteriota"="#FFFFB3",
    "Bacteroidota"="#BEBADA",
    "Firmicutes"= "#80B1D3",
    "Gemmatimonadota"= "#FDB462",
    "Myxococcota"="#FB8072",
    "Planctomycetota" = "#B3DE69",
    "Proteobacteria" = "#FCCDE5",
    "Verrucomicrobiota" = "#BC80BD",
    "less than 2.5%"="grey50",
    "not assigned" = "grey90")

```


```{r}
## calculate numbers for the paper
molten_dat_phylum_merge_16S %>% 
  group_by(Phylum)%>% 
  summarise(mean_value = mean(Abundance)) %>%
  arrange(desc(mean_value))
```


```{r}


p_phyla_16S <- ggplot(molten_dat_phylum_merge_16S, 
                  aes(x = code, 
                             y = Abundance,
                             fill = Taxonomy)) + 
  geom_bar(stat = "identity", position = "stack")+
  facet_grid(habitat  ~ management, scales = "free_x", space = "free_x")+
   scale_fill_manual(values = plotColorBarplot_16S)+
  labs( title = "Bacteria",
        y = "Relative abundance (%)", 
        x = NULL)+
  theme(legend.position="bottom")+
  theme(axis.text.x = element_text(size = 6))
  

p_phyla_16S


```


\newpage


# alpha diversity option A (with all the data)


## 16S calculate diversity indices

```{r, echo= FALSE}

otu_16S <- data.frame(otu_table(physeq_16S))

sample_16S <- data.frame(sample_data(physeq_16S))

sample_16S <- sample_16S %>% dplyr::select(c("traitement","habitat","code"))

richness <- as.data.frame(specnumber(otu_16S, MARGIN = 2)) 

colnames(richness) <- "richness"

# Jost Diversity
shannon <- as.data.frame(exp(diversity(otu_16S, MARGIN = 2))) # exponential of Shannon
                           
colnames(shannon) <- "shannon"

# number of sequences per sample
number_sequences_perSample <-as.data.frame(sample_sums(physeq_16S))

colnames(number_sequences_perSample) <- "number_sequences_perSample"


diversity_physeq_16S <-left_join(sample_16S%>% rownames_to_column("sample_names"), 
          richness%>% rownames_to_column("sample_names"),
          by = join_by(sample_names))

diversity_physeq_16S <-left_join(diversity_physeq_16S, 
          shannon%>% rownames_to_column("sample_names"),
          by = join_by(sample_names))


diversity_physeq_16S <-left_join(diversity_physeq_16S, 
          number_sequences_perSample%>% rownames_to_column("sample_names"),
          by = join_by(sample_names))

# Hill's ratio  also called Sheldon's evenness
diversity_physeq_16S <- diversity_physeq_16S  %>%
  mutate (evenness = shannon/richness)


```



## ITS calculate the diversity indices



```{r, echo= FALSE}

otu_ITS <- data.frame(otu_table(physeq_ITS))

sample_ITS <- data.frame(sample_data(physeq_ITS))

sample_ITS <- sample_ITS %>% dplyr::select(c("traitement","habitat","code"))

richness <- as.data.frame(specnumber(otu_ITS, MARGIN = 2)) 

colnames(richness) <- "richness"

# Jost Diversity
shannon <- as.data.frame(exp(diversity(otu_ITS, MARGIN = 2)))
                           
colnames(shannon) <- "shannon"

# number of sequences per sample
number_sequences_perSample <-as.data.frame(sample_sums(physeq_ITS))

colnames(number_sequences_perSample) <- "number_sequences_perSample"


diversity_physeq_ITS <-left_join(sample_ITS%>% rownames_to_column("sample_names"), 
          richness%>% rownames_to_column("sample_names"),
          by = join_by(sample_names))

diversity_physeq_ITS <-left_join(diversity_physeq_ITS, 
          shannon%>% rownames_to_column("sample_names"),
          by = join_by(sample_names))


diversity_physeq_ITS <-left_join(diversity_physeq_ITS, 
          number_sequences_perSample%>% rownames_to_column("sample_names"),
          by = join_by(sample_names))

# Hill's ratio  also called Sheldon's evenness
diversity_physeq_ITS <- diversity_physeq_ITS  %>%
  mutate (evenness = shannon/richness)


```

```{r}


# re-organize the data 
with_soil_16S <- diversity_physeq_16S |>
  mutate(traitement = if_else(is.na(traitement), "soil", traitement))  |>
  filter(traitement %in% c("ThermoSeed","control","synthetic","Tillecur", "soil")) |>
   mutate(traitement = factor(traitement, levels = c("soil", "control", "synthetic", "ThermoSeed","Tillecur"))) |>
  mutate(traitement = factor(traitement),              # convert to factor
         traitement = relevel(traitement, ref = "soil")) |>
   mutate(habitat = factor(habitat),              # convert to factor
         habitat = relevel(habitat, ref = "soil"))


with_soil_ITS <-  diversity_physeq_ITS|>
  mutate(traitement = if_else(is.na(traitement), "soil", traitement))  |>
  filter(traitement %in% c("ThermoSeed","control","synthetic","Tillecur", "soil")) |>
  mutate(traitement = factor(traitement, levels = c("soil", "control","synthetic", "ThermoSeed","Tillecur"))) |>
   mutate(habitat = factor(habitat),              # convert to factor
         habitat = relevel(habitat, ref = "soil"))



```


\newpage

## 16S Shannon (with soil for supplemental)

```{r}

# for Shannon, add as covariate the number of sequences per sample
model_16S_shannon <- lmer(log(shannon) ~ traitement +  log(number_sequences_perSample) +  (1 | code), data = with_soil_16S)

anova(model_16S_shannon)


#summary(model_16S_shannon)

```

```{r, eval =FALSE}
# Tukey-Anscomble
plot(y = residuals(model_16S_shannon, scaled = T), x = predict(model_16S_shannon), 
     main = "scaled residulas vs fitted", ylab = "scaled residuals", xlab = "fitted values")

## QQ-plot
qqnorm(residuals(model_16S_shannon), main = "QQ-plot of residuals")
qqline(residuals(model_16S_shannon))

```


```{r, echo = FALSE, eval = TRUE}
#levels(with_soil_16S$traitement)

contrast_res<- emmeans(model_16S_shannon,
        trt.vs.ctrl ~ traitement,
        ref = "soil",
        adjust = "dunnet")$contrasts

contrast_res

## add letters for the p.value < 0.05

res_df <- as.data.frame(summary(contrast_res))


letters_df <- res_df %>%
  mutate(
    traitement = str_remove(contrast, " - soil"),
    group = ifelse(p.value < 0.05, "b", "a")) %>%
  dplyr::select(traitement, group)
  
letters_df <- bind_rows(
  data.frame(traitement = "soil", group = "a"),
  letters_df
)

letters_df <- letters_df %>%
  mutate(habitat = ifelse(traitement =="soil", "soil","rhizosphere"))

#

y <- with_soil_16S$shannon

pos_df <- letters_df %>%
  group_by(traitement) %>%
  summarise(y_pos = max(y) + 0.1 * diff(range(y)))

plot_letters <- left_join(letters_df, pos_df, by = "traitement")


plot_letters <- plot_letters %>% 
  mutate(traitement = factor(traitement, 
                             levels = c("soil", 
                                        "control", 
                                        "synthetic",
                                        "ThermoSeed",
                                        "Tillecur")))%>% 
  mutate(habitat = factor(habitat, levels = c("soil", "rhizosphere")))
  
```

\newpage

## Plot 16S Shannon (all data)

```{r}



#summary(model_16S_all)


all_shannon_16S <- ggplot(with_soil_16S, 
                aes(x = traitement,
                   y = shannon)) + 
  geom_dotplot(aes(fill = traitement),
               binaxis = "y",
               stackdir = "center")+
   stat_summary(fun = "median", geom = "crossbar",  width = 0.3, color = "black") +
  geom_text(data = plot_letters, aes(x = traitement,
                                     y = y_pos, 
                                     label = group))+
  facet_grid(.~habitat, scales = "free", space = "free")+
    scale_fill_manual(values = custom_colors)+
      scale_x_discrete(guide = guide_axis(n.dodge = 2))+
  labs (title = "Bacteria",
        y = "Shannon" , 
        x =NULL)+
  my_theme+
     theme(legend.position = "none")

all_shannon_16S



```

\newpage

## 16S Shannon (only rhizo)


```{r}
rhizo_16S <- with_soil_16S %>% filter(habitat =="rhizosphere")%>% filter(traitement %in% c("ThermoSeed","control","synthetic","Tillecur"))

rhizo_16S$traitement <- droplevels(rhizo_16S$traitement)

```


```{r, echo = TRUE}

model_16S_shannon <- lmer(log(shannon) ~ traitement +  log(number_sequences_perSample) +  (1 | code), data = rhizo_16S)

anova(model_16S_shannon)


# summary(model_16S_log)
# 
# write.csv(summary(model_16S_log)$coefficient, "../results/16S_rhizo_log.csv")

#summary(model_16S_shannon)



```

```{r, eval = FALSE}
emmeans(model_16S_shannon,
        trt.vs.ctrl ~ traitement,
        ref = "control",
        adjust = "dunnet")$contrasts

```

```{r, eval = FALSE}

# Tukey-Anscomble

plot(y= residuals(model_16S_shannon, scaled = T), 
     x = predict(model_16S_shannon), 
     main = "scaled residulas vs fitted", ylab = "scaled residuals", xlab = "fitted values")

## QQ-plot
qqnorm(residuals(model_16S_shannon), 
       main = "QQ-plot of residuals")
qqline(residuals(model_16S_shannon))

#summary(model_16S_log)



```



```{r}


ctl <-rhizo_16S %>% 
  filter(traitement == "control") %>%
  dplyr::select(c("shannon","code","traitement"))
                   
 
treated <-rhizo_16S %>% 
  dplyr::filter(!traitement == "control") %>%
  dplyr::select(c("shannon","code","traitement"))

# make it wide to combine all alternative treatments
treated_wide <- treated %>%
  pivot_wider(names_from = traitement, values_from = shannon)


 
# make it long to join it with control
treated_long <- treated_wide %>% 
  pivot_longer(cols = synthetic:Tillecur,
               values_to = "shannon", 
               names_to = "traitement_name") 


for_gg <-right_join(ctl, treated_long, by = "code")
```


\newpage

## Plot log ratio Shannon (ITS all data)


```{r}

for_gg <- for_gg %>% 
  mutate (ratio = shannon.y/shannon.x)

for_gg <- for_gg %>%
  mutate(traitement_name = fct_relevel(traitement_name, "synthetic"))


# remove NA
for_gg <- for_gg %>%
  filter(!is.na(ratio))

shannon_16S <- ggplot(for_gg, 
                aes(x = traitement_name,
                   y = log(ratio)))+
  geom_dotplot(aes(fill = traitement_name),
               binaxis = "y",
               stackdir = "center")+
   stat_summary(fun = "median", geom = "crossbar",  width = 0.3, color = "black")+
     geom_hline(yintercept = 0, color = "red") +
  labs (title = "Bacteria",
        y = "log ratio Shannon" , 
        x =NULL)+
  scale_fill_manual(values = custom_colors)+
  my_theme+
  theme(legend.position = "none")

shannon_16S



```


\newpage


## ITS Shannon (with soil for supplemental)

```{r}

model_ITS_all <- lmer(shannon ~ traitement + log(number_sequences_perSample) + (1 | code), data = with_soil_ITS)

anova(model_ITS_all)



```

```{r, eval = FALSE}
# Tukey-Anscomble
plot(y = residuals(model_ITS_all, scaled = T), x = predict(model_ITS_all), 
     main = "scaled residulas vs fitted", ylab = "scaled residuals", xlab = "fitted values")


qqnorm(residuals(model_ITS_all), main = "QQ-plot of residuals")
qqline(residuals(model_ITS_all))
```


## Plot ITS Shannon (all data)

```{r}

all_shannon_ITS <- ggplot(with_soil_ITS, 
                aes(x = traitement,
                   y = shannon)) + 
  geom_dotplot(aes(fill = traitement),
               binaxis = "y",
               stackdir = "center")+
   stat_summary(fun = "median", geom = "crossbar",  width = 0.3, color = "black") +
  facet_grid(.~habitat, scales = "free", space = "free")+
    scale_fill_manual(values = custom_colors)+
    scale_x_discrete(guide = guide_axis(n.dodge = 2))+

  labs (title = "Fungi",
        y = "Shannon" , 
        x =NULL)+
  my_theme+
     theme(legend.position = "none")

all_shannon_ITS


```

\newpage




## ITS Shannon (only rhizo)


```{r}
rhizo_ITS <- with_soil_ITS %>% filter(habitat =="rhizosphere")%>% filter(traitement %in% c("ThermoSeed","control","synthetic","Tillecur"))

rhizo_ITS$traitement <- droplevels(rhizo_ITS$traitement)

```



```{r, echo = TRUE}



model_ITS_log <- lmer(log(shannon) ~ traitement+  log(number_sequences_perSample)+ (1 | code), 
                      data = rhizo_ITS )

anova(model_ITS_log)



```

```{r, eval = FALSE}

plot(y = residuals(model_ITS_log, scaled = T), 
     x = predict(model_ITS_log), 
     main = "scaled residulas vs fitted", ylab = "scaled residuals", xlab = "fitted values")

## QQ-plot
qqnorm(residuals(model_ITS_log), main = "QQ-plot of residuals")
qqline(residuals(model_ITS_log))

#summary(model1)



```


```{r}


ctl <-rhizo_ITS %>% 
  dplyr::filter(traitement == "control") %>%
  dplyr::select(c("shannon","code","traitement"))
                   
 
treated <-rhizo_ITS %>% 
  dplyr::filter(!traitement == "control") %>%
  dplyr::select(c("shannon","code","traitement"))

# make it wide to combine all alternative treatments
treated_wide <- treated %>%
  pivot_wider(names_from = traitement, values_from = shannon)


 
# meake it long to join it with control
treated_long <- treated_wide %>% 
  pivot_longer(cols = synthetic:Tillecur,
               values_to = "shannon", 
               names_to = "traitement_name") 


for_gg <-right_join(ctl, treated_long, by = "code")

```


## Plot log ratio Shannon (ITS all data)

```{r}

for_gg <- for_gg %>% 
  mutate (ratio = shannon.y/shannon.x)

 

# remove NA
for_gg <- for_gg %>%
  filter(!is.na(ratio))

ffor_gg <- for_gg %>%
  mutate(traitement_name = fct_relevel(traitement_name, "synthetic"))



shannon_ITS <- ggplot(for_gg, 
                aes(x = traitement_name,
                   y = log(ratio))) + 
  geom_dotplot(aes(fill = traitement_name),
               binaxis = "y",
               stackdir = "center")+
   stat_summary(fun = "median", geom = "crossbar",  width = 0.3, color = "black")+
  geom_hline(yintercept = 0, color = "red") +
  labs (title = "Fungi",
        y = "log ratio richness" , 
        x =NULL)+
  my_theme+
   scale_fill_manual(values = custom_colors)+
  theme(legend.position = "none")

shannon_ITS


```




# alpha diversity option B, rarefy data => supplemental

## Rarefy


```{r}
raremax_16S <- 9073



taxonTable_16S <-as.data.frame(otu_table(physeq_16S))



```


Bacterial communities are rarefied with `r raremax_16S`.

Fungal communities are rarefied with `r raremax_ITS`. 
Two samples are removed which are below the threshold.

```{r}

sample_names(subset_samples(physeq_ITS, colSums(otu_table(physeq_ITS)) < raremax_ITS))

# remove samples below threshold
physeq_ITS_rare <- subset_samples(physeq_ITS, colSums(otu_table(physeq_ITS)) > raremax_ITS)

# remove  OTUs with zero sequences
physeq_ITS_rare <-subset_taxa(physeq_ITS_rare, rowSums(otu_table(physeq_ITS_rare))>0 )


taxonTable_ITS_rare <-as.data.frame(otu_table(physeq_ITS_rare))

```



This code chunk is from Bodenhausen et al. 2019 

```{r rarefy_ITS, eval=FALSE, echo  = TRUE}

richness_mat <- c()
diversity_mat <- c()
even_mat <- c()

for (i in 1:500){
  # generate rarefied community
  
  temp_mat <- t(rrarefy(t(taxonTable_ITS_rare), sample = raremax_ITS))
  
  # richness is species number
  s <- specnumber(temp_mat, MARGIN=2)
  #cbind( colSums(temp_mat>0),specnumber(temp_mat, MARGIN=2))
  richness_mat <- cbind(richness_mat,s)
  
  # from Jost 2006: D=exp(H), exponential of shannon entropy
  shannon <- vegan::diversity(temp_mat, index="shannon", MARGIN=2)
  d <- exp(shannon)
  diversity_mat <- cbind(diversity_mat,d)
  
  # Hill's ratio  also called Sheldon's evenness
  j <- exp(shannon)/s
  even_mat <- cbind(even_mat,j)
}

# calculate means of subsamples
otu_diversity <- rowMeans(diversity_mat)
otu_richness <- rowMeans(richness_mat)
otu_evenness <- rowMeans(even_mat)

# put them all in one dataframe
summary_diversity_ITS <- data.frame( cbind( otu_diversity, otu_richness, otu_evenness))

sampleTable_rare <- sample_data(physeq_ITS_rare)

stopifnot(identical(rownames(sampleTable_rare), rownames(summary_diversity_ITS)))


for_anova_ITS <- cbind(summary_diversity_ITS, sampleTable_rare)

write.csv(for_anova_ITS, file="../final_results/for_anova_rarefied_ITS.csv")

```


```{r rarefy_16S, eval=FALSE}


# this code chunk is from my Petunia paper

richness_mat <- c()
diversity_mat <- c()
even_mat <- c()

for (i in 1:500){
  # generate rarefied community
  
  temp_mat <- t(rrarefy(t(taxonTable_16S), sample = raremax_16S))
  
  # richness is species number
  s <- specnumber(temp_mat, MARGIN=2)
  #cbind( colSums(temp_mat>0),specnumber(temp_mat, MARGIN=2))
  richness_mat <- cbind(richness_mat,s)
  
  # from Jost 2006: D=exp(H), exponential of shannon entropy
  shannon <- vegan::diversity(temp_mat, index="shannon", MARGIN=2)
  d <- exp(shannon)
  diversity_mat <- cbind(diversity_mat,d)
  
  # Hill's ratio  also called Sheldon's evenness
  j <- exp(shannon)/s
  even_mat <- cbind(even_mat,j)
}

# calculate means of subsamples
otu_diversity <- rowMeans(diversity_mat)
otu_richness <- rowMeans(richness_mat)
otu_evenness <- rowMeans(even_mat)

# put them all in one dataframe
summary_diversity_16S<- data.frame( cbind( otu_diversity, otu_richness, otu_evenness))

sampleTable_rare <- sample_data(physeq_16S)

stopifnot(identical(rownames(sampleTable_rare), rownames(summary_diversity_16S)))


for_anova_16S_rarefy <- cbind(summary_diversity_16S, sampleTable_rare)

write.csv(for_anova_16S, file="../final_results/for_anova_rarefied_16S.csv")

```





```{r}
for_anova_16S_rarefy <- read.csv("../final_results/for_anova_rarefied_16S.csv", header=TRUE)

for_anova_ITS_rarefy <- read.csv("../final_results/for_anova_rarefied_ITS.csv", header=TRUE)

# re-organize the data 
with_soil_16S_rarefy <- for_anova_16S_rarefy |>
  mutate(traitement = if_else(is.na(traitement), "soil", traitement))  |>
  filter(traitement %in% c("ThermoSeed","control","synthetic","Tillecur", "soil")) |>
   mutate(traitement = factor(traitement, levels = c("soil", "control", "synthetic", "ThermoSeed","Tillecur"))) |>
  mutate(traitement = factor(traitement),              # convert to factor
         traitement = relevel(traitement, ref = "soil")) |>
   mutate(habitat = factor(habitat),              # convert to factor
         habitat = relevel(habitat, ref = "soil"))


with_soil_ITS_rarefy <-  for_anova_ITS_rarefy|>
  mutate(traitement = if_else(is.na(traitement), "soil", traitement))  |>
  filter(traitement %in% c("ThermoSeed","control","synthetic","Tillecur", "soil")) |>
  mutate(traitement = factor(traitement, levels = c("soil", "control","synthetic", "ThermoSeed","Tillecur"))) |>
   mutate(habitat = factor(habitat),              # convert to factor
         habitat = relevel(habitat, ref = "soil"))



```



\newpage

## 16S Shannon (with soil for supplemental)

```{r}
model_16S_shannon <- lmer(otu_diversity ~ traitement + (1 | code), data = with_soil_16S_rarefy)

anova(model_16S_shannon)

#summary(model_16S_shannon)

```

```{r, eval =FALSE}
# Tukey-Anscomble
plot(y = residuals(model_16S_shannon, scaled = T), x = predict(model_16S_shannon), 
     main = "scaled residulas vs fitted", ylab = "scaled residuals", xlab = "fitted values")

## QQ-plot
qqnorm(residuals(model_16S_shannon), main = "QQ-plot of residuals")
qqline(residuals(model_16S_shannon))

```


```{r, echo = FALSE}
#levels(with_soil_16S$traitement)

contrast_res <- emmeans(model_16S_shannon,
        trt.vs.ctrl ~ traitement,
        ref = "soil",
        adjust = "dunnet")$contrasts

contrast_res



## add letters for the p.value < 0.05

res_df <- as.data.frame(summary(contrast_res))


letters_df <- res_df %>%
  mutate(
    traitement = str_remove(contrast, " - soil"),
    group = ifelse(p.value < 0.05, "b", "a")) %>%
  dplyr::select(traitement, group)
  
letters_df <- bind_rows(
  data.frame(traitement = "soil", group = "a"),
  letters_df
)

letters_df <- letters_df %>%
  mutate(habitat = ifelse(traitement =="soil", "soil","rhizosphere"))

#

y <- with_soil_16S_rarefy$otu_diversity

pos_df <- letters_df %>%
  group_by(traitement) %>%
  summarise(y_pos = max(y) + 0.1 * diff(range(y)))

plot_letters <- left_join(letters_df, pos_df, by = "traitement")


plot_letters <- plot_letters %>% 
  mutate(traitement = factor(traitement, 
                             levels = c("soil", 
                                        "control", 
                                        "synthetic",
                                        "ThermoSeed",
                                        "Tillecur")))%>% 
  mutate(habitat = factor(habitat, levels = c("soil", "rhizosphere")))
```

\newpage

## Plotting 16S Shannon (rarefied data)

```{r}



#summary(model_16S_all)


all_shannon_16S_rarefy <- ggplot(with_soil_16S_rarefy, 
                aes(x = traitement,
                   y = otu_diversity)) + 
  geom_dotplot(aes(fill = traitement),
               binaxis = "y",
               stackdir = "center")+
   stat_summary(fun = "median", geom = "crossbar",  width = 0.3, color = "black") +
  geom_text(data = plot_letters, aes(x = traitement,
                                     y = y_pos, 
                                     label = group))+
  facet_grid(.~habitat, scales = "free", space = "free")+
    scale_fill_manual(values = custom_colors)+
      scale_x_discrete(guide = guide_axis(n.dodge = 2))+
  labs (title = "Bacteria",
        y = "Shannon" , 
        x =NULL)+
  my_theme+
     theme(legend.position = "none")

all_shannon_16S_rarefy



```


\newpage

## only rhizo 16S Shannon rarefied


```{r}
rhizo_16S_rarefy <- with_soil_16S_rarefy %>% filter(habitat =="rhizosphere")%>% filter(traitement %in% c("ThermoSeed","control","synthetic","Tillecur"))

```


```{r, echo = TRUE}

model_16S_log <- lmer(log(otu_diversity) ~ traitement+ (1 | code), data = rhizo_16S_rarefy )

anova(model_16S_log)


```

```{r, eval = FALSE}

# Tukey-Anscomble

plot(y= residuals(model_16S_log, scaled = T), 
     x = predict(model_16S_log), 
     main = "scaled residulas vs fitted", ylab = "scaled residuals", xlab = "fitted values")

## QQ-plot
qqnorm(residuals(model_16S_log), 
       main = "QQ-plot of residuals")
qqline(residuals(model_16S_log))

#summary(model_16S_log)



```



```{r}


ctl <-rhizo_16S_rarefy %>% 
  filter(traitement == "control") %>%
  select(c("otu_diversity","code","traitement"))
                   
 
treated <-rhizo_16S_rarefy %>% 
  dplyr::filter(!traitement == "control") %>%
  select(c("otu_diversity","code","traitement"))

# make it wide to combine all alternative treatments
treated_wide <- treated %>%
  pivot_wider(names_from = traitement, values_from = otu_diversity)


 
# make it long to join it with control
treated_long <- treated_wide %>% 
  pivot_longer(cols = synthetic:Tillecur,
               values_to = "otu_diversity", 
               names_to = "traitement_name") 


for_gg <-right_join(ctl, treated_long, by = "code")
```


## Plot log ratio Shannon (16S rarefied)

```{r}

for_gg <- for_gg %>% 
  mutate (ratio = otu_diversity.y/otu_diversity.x)

for_gg <- for_gg %>%
  mutate(traitement_name = fct_relevel(traitement_name, "synthetic"))


# remove NA
for_gg <- for_gg %>%
  filter(!is.na(ratio))

rhizo_shannon_16S_rarefy <- ggplot(for_gg, 
                aes(x = traitement_name,
                   y = log(ratio)))+
  geom_dotplot(aes(fill = traitement_name),
               binaxis = "y",
               stackdir = "center")+
   stat_summary(fun = "median", geom = "crossbar",  width = 0.3, color = "black")+
     geom_hline(yintercept = 0, color = "red") +
  labs (title = "Bacteria",
        y = "log ratio Shannon" , 
        x =NULL)+
  scale_fill_manual(values = custom_colors)+
  my_theme+
  theme(legend.position = "none")

rhizo_shannon_16S_rarefy

```



\newpage


## ITS shannon (with soil for supplemental)

```{r}

model_ITS_all <- lmer(otu_diversity ~ traitement + (1 | code), data = with_soil_ITS_rarefy)

anova(model_ITS_all)

# m5 <- lmer(log(shannon) ~ traitement + (1 | code), data = rhizo)
# 
# anova(m5)


#summary(model_ITS_all)


```

```{r, eval = FALSE}
# Tukey-Anscomble
plot(y = residuals(model_ITS_all, scaled = T), x = predict(model_ITS_all), 
     main = "scaled residulas vs fitted", ylab = "scaled residuals", xlab = "fitted values")


qqnorm(residuals(model_ITS_all), main = "QQ-plot of residuals")
qqline(residuals(model_ITS_all))
```

## Plotting ITS Shannon (rarefied data)


```{r}

all_shannon_ITS_rarefy <- ggplot(with_soil_ITS_rarefy, 
                aes(x = traitement,
                   y = otu_diversity)) + 
  geom_dotplot(aes(fill = traitement),
               binaxis = "y",
               stackdir = "center")+
   stat_summary(fun = "median", geom = "crossbar",  width = 0.3, color = "black") +
  facet_grid(.~habitat, scales = "free", space = "free")+
    scale_fill_manual(values = custom_colors)+
    scale_x_discrete(guide = guide_axis(n.dodge = 2))+

  labs (title = "Fungi",
        y = "Shannon" , 
        x =NULL)+
  my_theme+
     theme(legend.position = "none")

all_shannon_ITS_rarefy


```

\newpage



## ITS Shannon (only rhizo)



```{r}
rhizo_ITS_rarefy <- with_soil_ITS_rarefy %>% filter(habitat =="rhizosphere")%>% filter(traitement %in% c("ThermoSeed","control","synthetic","Tillecur"))

```


```{r, echo = TRUE}

model_ITS_log <- lmer(log(otu_diversity) ~ traitement+ (1 | code), data = rhizo_ITS_rarefy )

anova(model_ITS_log)



```

```{r, eval = FALSE}

plot(y = residuals(model_ITS_log, scaled = T), 
     x = predict(model_ITS_log), 
     main = "scaled residulas vs fitted", ylab = "scaled residuals", xlab = "fitted values")

## QQ-plot
qqnorm(residuals(model_ITS_log), main = "QQ-plot of residuals")
qqline(residuals(model_ITS_log))

#summary(model1)



```


```{r}


ctl <-rhizo_ITS_rarefy %>% 
  filter(traitement == "control") %>%
  select(c("otu_diversity","code","traitement"))
                   
 
treated <-rhizo_ITS_rarefy %>% 
  dplyr::filter(!traitement == "control") %>%
  select(c("otu_diversity","code","traitement"))

# make it wide to combine all alternative treatments
treated_wide <- treated %>%
  pivot_wider(names_from = traitement, values_from = otu_diversity)


 
# meake it long to join it with control
treated_long <- treated_wide %>% 
  pivot_longer(cols = synthetic:Tillecur,
               values_to = "otu_diversity", 
               names_to = "traitement_name") 


for_gg <-right_join(ctl, treated_long, by = "code")

```

## Plot log ratio Shannon (ITS rarefied)


```{r}

for_gg <- for_gg %>% 
  mutate (ratio = otu_diversity.y/otu_diversity.x)

 

# remove NA
for_gg <- for_gg %>%
  filter(!is.na(ratio))

for_gg <- for_gg %>%
  mutate(traitement_name = fct_relevel(traitement_name, "synthetic"))



rhizo_shannon_ITS_rarefy <- ggplot(for_gg, 
                aes(x = traitement_name,
                   y = log(ratio))) + 
  geom_dotplot(aes(fill = traitement_name),
               binaxis = "y",
               stackdir = "center")+
   stat_summary(fun = "median", geom = "crossbar",  width = 0.3, color = "black")+
  geom_hline(yintercept = 0, color = "red") +
  labs (title = "Fungi",
        y = "log ratio richness" , 
        x =NULL)+
  my_theme+
   scale_fill_manual(values = custom_colors)+
  theme(legend.position = "none")

rhizo_shannon_ITS_rarefy


```





\newpage



# Beta diversity


## ITS PERMANOVA



```{r, echo =TRUE}

env_ITS <- data.frame(sample_data(physeq_ITS_norm))

otu_ITS <- vegdist(t(otu_table(physeq_ITS_norm, taxa_are_rows = FALSE)))

bd <- betadisper(otu_ITS, group = env_ITS$code)

anova(bd)

permutest(bd, permutations = 999)
# No evidence that groups differ in dispersion.
# The PERMANOVA assumption of homogeneity of dispersion is met.

adonis2(otu_ITS ~ code, data =  env_ITS)

adonis2(otu_ITS ~ habitat, data =  env_ITS)

adonis2(otu_ITS ~ date, data =  env_ITS)

adonis2(otu_ITS ~ management, data =  env_ITS)



```

 We observe a strong effect of `code`, `habitat`, and `date`, no effect for `management`.
 
In general, the four replicates from each farm are close to each other which is a good sign: there was no mixup in the samples and everything went well during the DNA extractions, PCR, and barcoding.

The first axis seperates the fields according to some gradient but the second axis seperates CH13 from the others.

 Very similar pattern with PCA made with Hellinger-transformed data and Euclidean distance.

 
\newpage

## ITS PCA


```{r PCA_ITS, eval = TRUE, fig.height = 7}

physeq_PCA <- ordinate(physeq_ITS_hell, "RDA", distance="euclidean")

# for the PCoA plots, nicer to use Sample_ID because they include site names
pca_ITS <- plot_ordination(physeq_ITS_hell,
                       physeq_PCA,axes=c(1,2),
                        color="code", 
                       shape="habitat",
                       label="code")+
                      geom_point(alpha=0.6)+
  my_theme+
  labs(title = "Fungi")+
    theme(legend.position = "none")

 pca_ITS
 

```

\newpage


\newpage

## 16S PERMANOVA

```{r, echo =TRUE}

env_16S <- data.frame(sample_data(physeq_16S_norm))

otu_16S <- vegdist(t(otu_table(physeq_16S_norm, taxa_are_rows = FALSE)))

bd <- betadisper(otu_16S, group = env_16S$code)

anova(bd)

permutest(bd, permutations = 999)


adonis2(otu_16S ~ code, data =  env_16S)

adonis2(otu_16S ~ habitat, data =  env_16S)

adonis2(otu_16S ~ date, data =  env_16S)

adonis2(otu_16S ~ management, data =  env_16S)



```

 We observe a strong effect of `code`, `habitat`, and `date`, only weak effect for `management` effect.
 
In general, the four replicates from each farm are close to each other which is a good sign: there was no mixup in the samples and everything went well during the DNA extractions, PCR, and barcoding.

The first axis seperates the fields according to some gradient but the second axis seperates CH13 from the others.

 Very similar pattern with PCA made with Hellinger-transformed data and Euclidean distance.

 
\newpage

## 16S PCA


```{r PCA_16S, eval = TRUE, fig.height = 7}

physeq_PCA <- ordinate(physeq_16S_hell, "RDA", distance="euclidean")

# for the PCoA plots, nicer to use Sample_ID because they include site names
pca_16S <- plot_ordination(physeq_16S_hell,
                       physeq_PCA,axes=c(1,2),
                        color="code", 
                       shape="habitat",
                       label="code")+
                      geom_point(alpha=0.6)+
  my_theme+
  labs(title = "Bacteria")+
    theme(legend.position = "none")

 pca_16S
 

```

\newpage


\newpage

# Constrained Analysis



## ITS add soil data

Subset only the non-treated samples


```{r}
physeq_ITS_untreated <- subset_samples(physeq_ITS, traitement =="control")

```


```{r}


soil_data <- read.csv("../final_data/TableS4.csv")

meta_data <- data.frame(sample_data(physeq_ITS_untreated)) 



## add soil data

with_soil <-left_join(meta_data %>% rownames_to_column("phyloseq_id"),
          soil_data, 
          by = "code")


## add data to phyloseq object
sample_data(physeq_ITS_untreated) <- with_soil%>%column_to_rownames("phyloseq_id")

```


```{r}
#  we like to have percentages
physeq_ITS_untreated_norm<-  transform_sample_counts(physeq_ITS_untreated, function(x) x / sum(x)*100 )

#colSums(otu_table(physeq_ITS_untreated_norm))

```

## prepare the data for RDA


```{r}
# OTU
taxonTable_ITS <-otu_table(physeq_ITS_untreated_norm, taxa_are_rows = FALSE)

taxonTable_ITS_hell <- sqrt(taxonTable_ITS) # hellinger is sqrt relative abundance

taxonTable_ITS_hell <- t(taxonTable_ITS_hell) # in vegan, samples are in the rows

env_data_ITS <-data.frame(sample_data(physeq_ITS_untreated_norm))

# small_env<- env_data_ITS %>% dplyr::select(-field,- library, - year)

small_env_ITS<- env_data_ITS %>% 
  dplyr::select(PH.H2O:Calcium_pourcent) %>% 
  select(-silt_pourcent)

# change names

small_env_ITS <- small_env_ITS %>%
  mutate(pH = PH.H2O,
         Corg = Corg_pourcent,
         clay = argile_pourcent,
         sand = sable_pourcent,
         P = phosphore_indice,
         K = potassium_indice,
         Mg = magnesium_indice,
         Ca = Calcium_pourcent,
                     .keep ="unused")


```


## ITS forward selection with ordistep

```{r, echo=FALSE}

# with 0 there is no constraint => PCA

mod0_ITS<- rda(taxonTable_ITS_hell ~  1, data = small_env_ITS) # Model with intercept only

mod1_ITS <- rda(taxonTable_ITS_hell ~ ., data = small_env_ITS) # Model with all explanatory variables

anova(mod1_ITS)
```


```{r, echo = FALSE, eval = TRUE}
# ordiR2step takes a bit less time than ordistep
# only run this code chunk by hand

(adjR2.tbrda <- RsquareAdj(mod1_ITS)$adj.r.squared)
#  0.1950007 - adjusted R2 explained by all variables is 19%

# 
# mod3_ITS<- ordiR2step (mod0_ITS, scope = formula (mod1_ITS), R2scope = adjR2.tbrda, direction = 'forward', permutations = 999)
# # 
 # save(mod3_ITS, file="../final_results/ordi2step_ITS.rda")
# 
load("../final_results/ordi2step_ITS.rda")

mod3_ITS$anova

mod3_ITS$call


write.csv(mod3_ITS$anova, "../final_results/TableS10B.csv")




RsquareAdj(mod3_ITS)$adj.r.squared
```

\newpage

## ITS Triplot RDA 



```{r prepare_data_triplot_ITS}

#ord$eig
eig2 <- eigenvals(mod3_ITS)
#eig2 / sum(eig2)
#cumsum(eig2 / sum(eig2))

percent_RDA <-round(eig2 / sum(eig2)*100,1)

coordi_RDA <- (paste("RDA", c(1:9),
                 paste(" (",percent_RDA[1:9],"%",")",sep=""),
                 sep=""))


species_scores_rda <-scores(mod3_ITS, 
                 choice=1:2,
                 scaling="species",
                 display="sp")

sites_scores_rda <-scores(mod3_ITS, 
                 choice=1:2,
                 scaling="species",
                 display="lc")




env_arrow_rda <- mod3_ITS$CCA$biplot[,1:2]

# Settings
ahead <- 0.05   # Length of arrow heads
offset <- 0.1 # distance between arrow head and center of blue text
mult_arrow <- 3
mult_species <- 2.5

on_the_left  <-sites_scores_rda %>% 
  as.data.frame() %>%
  tibble::rownames_to_column("sample_name")

on_the_right <-env_data_ITS %>% 
  rownames_to_column("sample_name")%>%
  dplyr::select(code, sample_name)

sites <- left_join(on_the_left, 
                   on_the_right , 
                   by = "sample_name") 

sites <- sites %>%
  mutate(code = factor(code, levels = sort(unique(as.character(code)))))

#levels(sites$code)


species <- data.frame()

# p 219 Borcard
spe.good.rda <- goodness(mod3_ITS) # by default species
sel.sp.rda <- which (spe.good.rda[,2]>=0.3) # cumulative


species <-  (species_scores_rda * mult_species) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("OTU") %>% 
  #mutate(has_label = OTU %in% top_OTU$otu)
  mutate(has_label = OTU %in% names(sel.sp.rda))

# alternatively label extreme sites
species_label <- species %>% 
  filter(abs(RDA1)>1 | abs(RDA2)>1)



environment <- (env_arrow_rda * mult_arrow) %>%
  as.data.frame() %>% 
  dplyr::select("RDA1", "RDA2") %>%
  tibble::rownames_to_column("environment") %>%
  mutate(len = sqrt(RDA1^2 + RDA2^2), 
       hoffset = RDA1/len*offset, voffset = RDA2/len*offset)

```



```{r plot_triplot_ITS}

# Plot
rda_ITS <- ggplot() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  # geom_point(aes(RDA1, RDA2), color = "grey", 
  #            shape = "cross", data = species) +
  # geom_text(aes(RDA1, RDA2, label = OTU), color = "grey", 
  #             hjust = 0, nudge_x = .05, 
  #           data = species %>% filter(.data$has_label)) +
  geom_point(aes(RDA1, RDA2, color = code), data = sites) +
  geom_text(aes(RDA1, RDA2, color = code, label = code), 
              hjust = 0, nudge_x = .05, data = sites) +
  geom_segment(aes(x = 0, y = 0, xend = RDA1, yend = RDA2), 
                 color = "blue", 
               arrow = arrow(length=unit(0.02,"npc"), type = "open"),
               data = environment) +
  geom_text(aes(RDA1 + hoffset, RDA2 + voffset, 
                label = environment), color = "blue",
            data = environment) +
  coord_equal(ratio = 1, clip = "off") +
  theme_bw() +
  labs(title = "Fungi",
       x = coordi_RDA[1], 
       y = coordi_RDA[2]) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(legend.position = "none")

print(rda_ITS) 
  

```

\newpage


## 16S add soil data

Subset only the non-treated samples


```{r}
physeq_16S_untreated <- subset_samples(physeq_16S, traitement =="control")



# decide to go back to Excel and try to make the identifier in excel instead of R

meta_data <- data.frame(sample_data(physeq_16S_untreated)) 



## add soil data

with_soil <-left_join(meta_data %>% rownames_to_column("phyloseq_id"),
          soil_data, 
          by =c("code"))


## add data to phyloseq object
sample_data(physeq_16S_untreated) <- with_soil%>%column_to_rownames("phyloseq_id")

```


```{r}
#  we like to have percentages
physeq_16S_untreated_norm<-  transform_sample_counts(physeq_16S_untreated, function(x) x / sum(x)*100 )

# colSums(otu_table(physeq_16S_untreated_norm))


```

## prepare data for RDA


```{r}
# OTU
taxonTable_16S <-otu_table(physeq_16S_untreated_norm, taxa_are_rows = FALSE)

taxonTable_16S_hell <- sqrt(taxonTable_16S) # hellinger is sqrt relative abundance

taxonTable_16S_hell <- t(taxonTable_16S_hell) # in vegan, samples are in the rows

env_data_16S <-data.frame(sample_data(physeq_16S_untreated_norm))

# small_env_16S<- env_data_16S %>% dplyr::select(-field,- library, - year)

small_env_16S<- env_data_16S %>% 
  dplyr::select(PH.H2O:Calcium_pourcent)%>% 
  select(-silt_pourcent)

# change names

small_env_16S <- small_env_16S %>%
  mutate(pH = PH.H2O,
         Corg = Corg_pourcent,
         clay = argile_pourcent,
         sand = sable_pourcent,
         P = phosphore_indice,
         K = potassium_indice,
         Mg = magnesium_indice,
         Ca = Calcium_pourcent,
                     .keep ="unused")


```


## 16S forward selection with ordistep

```{r, echo=FALSE}

# with 0 there is no constraint => PCA

mod0_16S<- rda(taxonTable_16S_hell ~  1, data = small_env_16S)

mod1_16S <- rda(taxonTable_16S_hell ~ ., data = small_env_16S)

anova(mod1_16S)
```


```{r, echo = FALSE, eval = TRUE}
# ordiR2step takes a bit less time than ordistep
# only run this code chunk by hand

(adjR2.tbrda <- RsquareAdj(mod1_16S)$adj.r.squared)
#   0.3156094 - adjusted R2 explained by all variables is 32%



# 
# mod3_16S<- ordiR2step (mod0_16S, scope = formula (mod1_16S), R2scope = adjR2.tbrda, direction = 'forward', permutations = 999)
# #    
# save(mod3_16S, file="../final_results/ordi2step_16S.rda")

load("../final_results/ordi2step_16S.rda")

mod3_16S$anova
mod3_16S$call


write.csv(mod3_16S$anova, "../final_results/TableS10A.csv")



RsquareAdj(mod3_16S)$adj.r.squared
```



\newpage

## 16S Triplot RDA 



```{r prepare_data_triplot_16S}

#ord$eig
eig2 <- eigenvals(mod3_16S)
#eig2 / sum(eig2)
#cumsum(eig2 / sum(eig2))

percent_RDA <-round(eig2 / sum(eig2)*100,1)

coordi_RDA <- (paste("RDA", c(1:9),
                 paste(" (",percent_RDA[1:9],"%",")",sep=""),
                 sep=""))


species_scores_rda <-scores(mod3_16S, 
                 choice=1:2,
                 scaling="species",
                 display="sp")

sites_scores_rda <-scores(mod3_16S, 
                 choice=1:2,
                 scaling="species",
                 display="lc")




env_arrow_rda <- mod3_16S$CCA$biplot[,1:2]


# Settings
ahead <- 0.05   # Length of arrow heads
offset <- 0.1 # distance between arrow head and center of blue text
mult_arrow <- 3
mult_species <- 2.5

on_the_left  <-sites_scores_rda %>% 
  as.data.frame() %>%
  tibble::rownames_to_column("sample")

on_the_right <-env_data_16S %>% 
  dplyr::select(code)%>%
  tibble::rownames_to_column("sample")

sites <- left_join(on_the_left, 
                   on_the_right , 
                   by = "sample")

sites <- sites %>%
  mutate(code = factor(code, levels = sort(unique(as.character(code)))))


species <- data.frame()

# p 219 Borcard
spe.good.rda <- goodness(mod3_16S) # by default species
sel.sp.rda <- which (spe.good.rda[,2]>=0.3) # cumulative


species <-  (species_scores_rda * mult_species) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("OTU") %>% 
  #mutate(has_label = OTU %in% top_OTU$otu)
  mutate(has_label = OTU %in% names(sel.sp.rda))

# alternatively label extreme sites
species_label <- species %>% 
  filter(abs(RDA1)>1 | abs(RDA2)>1)



environment <- (env_arrow_rda * mult_arrow) %>%
  as.data.frame() %>% 
  dplyr::select("RDA1", "RDA2") %>%
  tibble::rownames_to_column("environment") %>%
  mutate(len = sqrt(RDA1^2 + RDA2^2), 
       hoffset = RDA1/len*offset, voffset = RDA2/len*offset)

```



```{r plot_triplot_16S}

# Plot
rda_16S <- ggplot() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  # geom_point(aes(RDA1, RDA2), color = "grey", 
  #            shape = "cross", data = species) +
  # geom_text(aes(RDA1, RDA2, label = OTU), color = "grey", 
  #             hjust = 0, nudge_x = .05, 
  #           data = species %>% filter(.data$has_label)) +
  geom_point(aes(RDA1, RDA2, color = code), data = sites) +
  geom_text(aes(RDA1, RDA2, color = code, label = code), 
              hjust = 0, nudge_x = .05, data = sites) +
  geom_segment(aes(x = 0, y = 0, xend = RDA1, yend = RDA2), 
                 color = "blue", 
               arrow = arrow(length=unit(0.02,"npc"), type = "open"),
               data = environment) +
  geom_text(aes(RDA1 + hoffset, RDA2 + voffset, 
                label = environment), color = "blue",
            data = environment) +
  coord_equal(ratio = 1, clip = "off") +
  theme_bw() +
  labs(title = "Bacteria",
       x = coordi_RDA[1], 
       y = coordi_RDA[2]) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(legend.position = "none")

print(rda_16S) 
  


```


\newpage

# Figures with cowplot


##  Figure 4

```{r}

two_alpha <-plot_grid(rhizo_shannon_16S_rarefy +theme_main,
                      rhizo_shannon_ITS_rarefy +theme_main, 
                         labels = c('A', 'B'),
                      ncol = 2,
                      label_size = 12)

#print(two_alpha)

if(makeFig == TRUE) ggsave("../final_figures/Figure4.png",
                           two_alpha,
                           width = 17, 
                           height = 8.5, 
                           units = "cm")




```


## Figure 5

```{r}


two_rda <-plot_grid(rda_16S + theme_main,
                      rda_ITS + theme_main, 
                         labels = c('A', 'B'),
                      ncol = 2,
                      label_size = 12)

#print(two_rda)


if(makeFig == TRUE) ggsave("../final_figures/Figure5.png",
                           two_rda,
                           width = 17, 
                           height = 8.5, 
                           units = "cm")

```

## Figure S3

```{r}

two_rare <-plot_grid(rare_16S + theme_main,
                      rare_ITS + theme_main, 
                         labels = c('A', 'B'),
                      nrow = 2,
                      label_size = 12)

#print(two_rare)

if(makeFig == TRUE) ggsave("../final_figures/FigureS3.png",
                           two_rare,
                           width = 16, 
                           height = 18, 
                           units = "cm")


```



## Figure S4

```{r}

two_stacked <-plot_grid(p_phyla_16S+theme_main ,
                      p_phyla_ITS , 
                         labels = c('A', 'B'),
                      nrow = 2,
                      label_size = 12)

#print(two_stacked)

if(makeFig == TRUE) ggsave("../final_figures/FigureS4.png",
                           two_stacked,
                           width = 16, 
                           height = 18, 
                           units = "cm")


```

##  Figure S5 C and D

```{r}

two_alpha <-plot_grid(all_shannon_16S +theme_main,
                      all_shannon_ITS +theme_main, 
                         labels = c('A', 'B'),
                      ncol = 2,
                      label_size = 12)

#print(two_alpha)

if(makeFig == TRUE) ggsave("../final_figures/FigureS5AB.png",
                           two_alpha,
                           width = 17, 
                           height = 8.5, 
                           units = "cm")



```


##  Figure S5 C and D

```{r}

two_alpha <-plot_grid(all_shannon_16S_rarefy +theme_main,
                      all_shannon_ITS_rarefy +theme_main, 
                         labels = c('C', 'D'),
                      ncol = 2,
                      label_size = 12)

#print(two_alpha)

if(makeFig == TRUE) ggsave("../final_figures/FigureS5CD.png",
                           two_alpha,
                           width = 17, 
                           height = 8.5, 
                           units = "cm")



```


##  Figure S6

```{r}

two_alpha <-plot_grid(rhizo_shannon_16S_rarefy +theme_main,
                      rhizo_shannon_ITS_rarefy +theme_main, 
                         labels = c('A', 'B'),
                      ncol = 2,
                      label_size = 12)

#print(two_alpha)

if(makeFig == TRUE) ggsave("../final_figures/FigureS6.png",
                           two_alpha,
                           width = 17, 
                           height = 8.5, 
                           units = "cm")




```

## Figure S7

```{r}


two_pca <-plot_grid(pca_16S + theme_main,
                      pca_ITS + theme_main, 
                         labels = c('A', 'B'),
                      ncol = 2,
                      label_size = 12)

# print(two_pca)


if(makeFig == TRUE) ggsave("../final_figures/FigureS7.png",
                           two_pca,
                           width = 17, 
                           height = 8.5, 
                           units = "cm")


```

## to do

- add the non-rarefied data analysis (Figure 2 and Figur S5 AB)
 
- make sure to only save data which is actually present in the MS (supplementary tables)
